### 2022-06-14 manuscript.Rmd
Small changes for new submission

author:
  - name          : "Alexander Enge"
    affiliation   : "1,2,*"
  - name          : "Shreya Kapoor"
    affiliation   : "1"
  - name          : "Anne-Sophie Kieslinger"
    affiliation   : "1"
  - name          : "Michael A. Skeide"
    affiliation   : "1"
affiliation:
  - id            : "1"
    institution   : "Research Group Learning in Early Childhood, Max Planck Institute for Human Cognitive and Brain Sciences, Stephanstraße 1a, Leipzig, Germany"
  - id            : "2"
    institution   : "Department of Psychology, Humboldt-Universität zu Berlin, Rudower Chaussee 18, Berlin, Germany"
  - id            : "*"
    institution   : "Corresponding author (enge@cbs.mpg.de)"

keywords          : "mental rotation, spatial cognition, spatial ability, infants, children, gender, meta-analysis"

bibliography      : "manuscript_files/references.bib"
csl               : "manuscript_files/nature.csl"

@@

Mental rotation, the cognitive process of moving an object in mind to predict how it looks in a new orientation, is tightly coupled to intelligence, learning, and educational achievement [@shepard1971; @hegarty1999; @johnson2005]. On average, males solve mental rotation tasks slightly faster than females [@linn1985; @voyer1995; @voyer2011; @maeda2013]. When such behavioral differences emerge during development, however, remains poorly understood [@lauer2019a]. Here we analyzed effect sizes derived from 59 experiments conducted in 1,798 infants aged 3--16 months. We robustly found that male infants recognized rotated objects slightly more reliably than female infants. These findings indicate that subtle gender differences in mental rotation are already present in the first months of life.

# Main

@@

In the present study, we meta-analyzed `r descriptives$n_experiments` effect sizes derived from looking times in mental rotation tasks performed by `r formatC(descriptives$n_infants, big.mark = ",")` infants (`r print_perc(descriptives$percent_female)` female) aged `r descriptives$min_age_months` to `r descriptives$max_age_months` months (mean age of `r descriptives$mean_age_weighted`; Supplementary Table\ \@ref(tab:tabs1))[@kellman1984; @rochat1996; @hespos1997; @moore2008; @quinn2008; @moore2011; @frick2013b; @mohring2013; @schwarzer2013; @schwarzer2013a; @frick2014b; @quinn2014; @lauer2015; @antrilli2016; @christodoulou2016; @lauer2016; @constantinescu2018; @erdmann2018; @gerhard2018; @slone2018; @kaaz2020; @gerhard-samunda2021; @kelch2021]. All tasks were embedded either into habituation experiments (`r descriptives$n_experiments_habituation`) or violation of expectation experiments (`r descriptives$n_experiments_voe`) [@fantz1964; @baillargeon1985]. These experiments comprised real world stimuli (e.g., toy objects) [@antrilli2016], three-dimensional digital stimuli (e.g., cube figures) [@moore2008], or two-dimensional digital stimuli (e.g., digits) [@quinn2008]. In habituation experiments, infants repeatedly saw an object until their looking times declined before they were presented with a mirror image of the object or with the familiar object at a new angle. Longer looking times at the mirror image were taken as evidence that an infant still recognized the familiar object after rotation through the new orientation. In violation of expectation experiments, infants were habituated to an object that was revolving repeatedly through a certain angle. Then, either the familiar object or an unseen object was shown while they were revolving through a certain angle. Subsequently, the object disappeared behind an occluder. Finally, the occluder was removed and either the object or its mirror object were shown. Larger differences in looking times for the familiar object versus the unseen mirror image (at the new angle) were taken as evidence that an infant still recognized the familiar object after rotation through the new angle. In contrast, looking times for both objects were expected to be similar if an infant did not recognize the familiar object from the new angle.

\vspace{1cm}

@@

```{r, fig1, include=TRUE, fig.height=14, fig.pos="bh", fig.width=12, fig.cap="(ref:fig1-caption)"}
# Get posterior draws for the effect *in each experiment*
epred_draws_meta <- epred_draws(res_meta, dat_meta, ndraws = NULL) %>%
  mutate(experiment_f = factor(experiment))

@@

@@

## Discussion

@@

The present study robustly revealed that male infants look slightly longer at novel rotated objects than female infants. Thus, on average, males show slightly more reliable mental rotation behavior already in the first months of life.

# Methods

@@

## Data availability

The datasets generated and analyzed during the current study are available in an Open Science Framework repository [@enge2022a], [https://osf.io/k3wdg/?view_only=06a4a0ac3d5f4681baab11751c1498b8](https://osf.io/k3wdg/?view_only=06a4a0ac3d5f4681baab11751c1498b8). This link is intended for peer review only and will be replaced with a public DOI link upon publication.

## Code availability

The analysis code for the current study is available in an Open Science Framework repository [@enge2022a], [https://osf.io/k3wdg/?view_only=06a4a0ac3d5f4681baab11751c1498b8](https://osf.io/k3wdg/?view_only=06a4a0ac3d5f4681baab11751c1498b8). This link is intended for peer review only and will be replaced with a public DOI link upon publication.

@@

# Author contributions

M.A.S. conceived, designed, and obtained funding for the study. A.E., S.K., and A.-S.K. (in consultation with M.A.S.) designed and executed the search and selection strategies, performed data analysis, conducted robustness checks, and visualized the results. A.E. and M.A.S. wrote the manuscript with additional input from S.K. and A.-S.K.

# Competing interests

The authors declare no competing interests.

# Materials & Correspondence

Supplementary Information is available for this paper. Correspondence and requests for materials should be addressed to Alexander Enge ([enge@cbs.mpg.de](mailto:enge@cbs.mpg.de)).

>>>>

author:
  - name          : "Alexander Enge"
    affiliation   : "1,2,*"
  - name          : "Shreya Kapoor"
    affiliation   : "1"
  - name          : "Anne-Sophie Kieslinger"
    affiliation   : "1"
  - name          : "Michael A. Skeide"
    affiliation   : "1"
affiliation:
  - id            : "1"
    institution   : "Research Group Learning in Early Childhood, Max Planck Institute for Human Cognitive and Brain Sciences, Stephanstraße 1a, 04103 Leipzig, Germany"
  - id            : "2"
    institution   : "Department of Psychology, Humboldt-Universität zu Berlin, Rudower Chaussee 18, 12489 Berlin, Germany"
  - id            : "*"
    institution   : "Corresponding author (enge@cbs.mpg.de)"

bibliography      : "manuscript_files/references.bib"
csl               : "manuscript_files/pnas.csl"

@@

Mental rotation, the cognitive process of moving an object in mind to predict how it looks in a new orientation, is tightly coupled to intelligence, learning, and educational achievement [@shepard1971; @hegarty1999; @johnson2005]. On average, males solve mental rotation tasks slightly faster than females [@linn1985; @voyer1995; @voyer2011; @maeda2013]. When such behavioral differences emerge during development, however, remains poorly understood [@lauer2019a]. Here we analyzed effect sizes derived from 59 experiments conducted in 1,798 infants aged 3--16 months. We robustly found that male infants recognized rotated objects more reliably than female infants. These findings indicate that gender differences in mental rotation are already present in the first months of life.

# Significance statement

Are there gender differences in cognitive ability? So far, putative differences in most cognitive domains turned out to be myths. However, mental rotation, the ability to move an object in mind to predict how it would look in a new orientation, is considered as a serious candidate. Here we investigated how early during development mental rotation behavior differs in male and female infants. We found that, on average, males recognized rotated objects more reliably than females as early as in the first months of life. These results are an important step towards narrowing down possible social and genetic origins of gender differences in mental rotation ability.

# Introduction

@@

In the present study, we meta-analyzed `r descriptives$n_experiments` effect sizes derived from looking times in mental rotation tasks performed by `r formatC(descriptives$n_infants, big.mark = ",")` infants (`r print_perc(descriptives$percent_female)` female) aged `r descriptives$min_age_months` to `r descriptives$max_age_months` months (mean age of `r descriptives$mean_age_weighted`; Supplementary Table\ \@ref(tab:tabs1))[@kellman1984; @rochat1996; @hespos1997; @moore2008; @quinn2008; @moore2011; @frick2013b; @mohring2013; @schwarzer2013; @schwarzer2013a; @frick2014b; @quinn2014; @lauer2015; @antrilli2016; @christodoulou2016; @lauer2016; @constantinescu2018; @erdmann2018; @gerhard2018; @slone2018; @kaaz2020; @gerhard-samunda2021; @kelch2021]. All tasks were embedded either into habituation experiments (`r descriptives$n_experiments_habituation` experiments) or violation of expectation experiments (`r descriptives$n_experiments_voe` experiments) [@fantz1964; @baillargeon1985]. These experiments comprised real world stimuli (e.g., toy objects), three-dimensional digital stimuli (e.g., cube figures), or two-dimensional digital stimuli (e.g., digits). In habituation experiments, infants repeatedly saw an object until their looking times declined before they were presented with a mirror image of the object or with the familiar object at a new angle. Longer looking times at the mirror image were taken as evidence that an infant still recognized the familiar object after rotation through the new orientation. In violation of expectation experiments, infants were habituated to an object that was revolving repeatedly through a certain angle. Then, either the familiar object or an unseen object was shown while they were revolving through a certain angle. Subsequently, the object disappeared behind an occluder. Finally, the occluder was removed and either the object or its mirror object were shown. Larger differences in looking times for the familiar object versus the unseen mirror image (at the new angle) were taken as evidence that an infant still recognized the familiar object after rotation through the new angle. In contrast, looking times for both objects were expected to be similar if an infant did not recognize the familiar object from the new angle.

@@

```{r, fig1, include=TRUE, fig.height=14, fig.width=12, fig.cap="(ref:fig1-caption)"}
# Get posterior draws for the effect *in each experiment*
epred_draws_meta <- epred_draws(res_meta, dat_meta, ndraws = NULL) %>%
  mutate(experiment_f = factor(experiment))

@@

# Results

@@

# Discussion

@@

The present study robustly revealed that male infants look longer at novel rotated objects than female infants. Thus, on average, males show more reliable mental rotation behavior already in the first months of life.

# Materials and methods

@@

## Data and code availability

The data and code for this study are available in an Open Science Framework repository ([https://doi.org/10.17605/OSF.IO/K3WDG](https://doi.org/10.17605/OSF.IO/K3WDG)).

@@
